Nix for Polyglot, Reproducible Data Science Workflows

### 2025-10-28 paper.qmd
Update paper.qmd

However, while Docker excels at providing this form of spatial reproducibility,
its application in interactive research workflows reveals several significant
limitations. First, the container model is often poorly aligned with the
iterative, interactive nature of data analysis. Running graphical applications
like RStudio Desktop requires complex and platform-specific configuration, while
ephemeral container filesystems mean that changes made during a session are lost
by default. Consequently, a common workflow involves developing an analysis
interactively on a host system and only containerizing it post-hoc, which
achieves reproducibility as an afterthought rather than as a continuous
guarantee during the research process. Packages like @dockerfiler are specifically
made to enable this type of post-hoc reproducibility.

@@

We strive to provide an available date per week: each Monday, a GitHub Action
tests popular [R]{.proglang} packages on Linux and macOS, and only if all tests
succeed is the date added to the list of available dates in [rix]{.pkg}. Users
can see all available dates with `rix::available_dates()`. This ensures users
can reliably install packages, and allows us to backport fixes if needed. For
example, when RStudio was temporarily broken due to a dependency issue
(`boost`), a pull request was submitted to the official `nixpkgs` repository. I
backported the fix to the `rstats-on-nix` fork, making RStudio available to
users of [rix]{.pkg} earlier than upstream, as merging PRs in the official
repository can take some time.

@@

The `rxp_populate()` function translates this [R]{.proglang} list into a
`pipeline.nix` file, which declaratively defines the entire workflow. Data flows
from derivation to derivation by being serialised into the efficient and
language-agnostic Arrow format using the pair of `encoder/decoder` functions.
Other universal formats, such as `csv` or `json` could have been used. It should
also be noted that if interoperability libraries such as [reticulate]{.pkg} or 
[rds2py]{.pkg} are available, it becomes possible to seamlessly transfer
supported objects to and from [R]{.proglang} and [Python]{.proglang} (for now,
such a feature is not yet integrated for conversion of arbitrary [Julia]{.proglang} objects).

>>>>

However, while Docker excels at providing this form of spatial reproducibility,
its application in interactive research workflows reveals several significant
limitations. First, the container model is often poorly aligned with the
iterative, interactive nature of data analysis. Running graphical applications
like RStudio Desktop requires complex and platform-specific configuration, while
ephemeral container filesystems mean that changes made during a session are lost
by default. Consequently, a common workflow involves developing an analysis
interactively on a host system and only containerizing it post-hoc, which
achieves reproducibility as an afterthought rather than as a continuous
guarantee during the research process. Packages like [dockerfiler]{.pkg} 
[@dockerfiler] are specifically made to enable this type of post-hoc reproducibility.

@@

We strive to provide an available date per week: each Monday, a GitHub Action
tests popular [R]{.proglang} packages on Linux and macOS, and only if all tests
succeed is the date added to the list of available dates in [rix]{.pkg}. Users
can see all available dates with `rix::available_dates()`. This ensures users
can reliably install packages, and allows us to backport fixes if needed. For
example, when RStudio was temporarily broken due to a dependency issue
([boost]{.pkg}), a pull request was submitted to the official `nixpkgs` repository. I
backported the fix to the `rstats-on-nix` fork, making RStudio available to
users of [rix]{.pkg} earlier than upstream, as merging PRs in the official
repository can take some time.

@@

The `rxp_populate()` function translates this [R]{.proglang} list into a
`pipeline.nix` file, which declaratively defines the entire workflow. Data flows
from derivation to derivation by being serialised into the efficient and
language-agnostic [Arrow]{.pkg} format using the pair of `encoder/decoder` functions.
Other universal formats, such as `csv` or `json` could have been used. It should
also be noted that if interoperability libraries such as [reticulate]{.pkg} or 
[rds2py]{.pkg} are available, it becomes possible to seamlessly transfer
supported objects to and from [R]{.proglang} and [Python]{.proglang} (for now,
such a feature is not yet integrated for conversion of arbitrary [Julia]{.proglang} objects).
